---
title: "Cross-column consistency: speed, travel distance, and floors"
format:
  html:
    embed-resources: true
execute:
  echo: true
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
---

```{r}
#| label: setup
#| message: false
library(dplyr)
library(ggplot2)
library(stringr)

elev <- nanoparquet::read_parquet("elevators.parquet")
```

The three columns `dv_speed_fpm`, `dv_travel_distance`, and the floor pair
`dv_floor_from`/`dv_floor_to` each describe the physical behaviour of a
device. They should be mutually consistent: an elevator's travel distance
should match the number of floors it serves, and its speed should imply a
reasonable travel time. This document checks those relationships and proposes
rules for setting implausible values to missing.

```{r}
#| label: coverage
elev |>
  mutate(
    speed_missing    = is.na(dv_speed_fpm),
    distance_missing = is.na(dv_travel_distance),
    floors_missing   = is.na(dv_floor_from) | is.na(dv_floor_to)
  ) |>
  count(speed_missing, distance_missing, floors_missing)
```

The most complete combination — all three fields present — covers about 37% of
records. Floor pairs are the most widely recorded field; travel distance is the
sparsest. The cross-checks below draw on whichever subset has both relevant
fields recorded.

## Speed

```{r}
#| label: speed-dist
#| fig-height: 3
elev |>
  filter(!is.na(dv_speed_fpm)) |>
  ggplot(aes(dv_speed_fpm)) +
  geom_histogram(binwidth = 25, boundary = 0) +
  scale_x_continuous(limits = c(0, 2100)) +
  labs(x = "Speed (fpm)", y = "Devices",
       caption = "Truncated at 2,100 fpm; 27 records above that limit not shown.")
```

The distribution is dominated by a handful of standard speeds (100, 200, 350
fpm for low-rise residential; 700–1,200 fpm for commercial high-rise). The
tails hold the problems.

### Too slow

```{r}
#| label: speed-slow
elev |>
  filter(!is.na(dv_speed_fpm), dv_speed_fpm < 10) |>
  count(dv_speed_fpm, device_type, sort = TRUE)
```

57 devices record a speed below 10 fpm, including 2 at exactly 0. The
fastest a person climbs stairs is roughly 60 fpm; no mechanically-driven
vertical transport device operates below ~10 fpm in normal service. Escalators
dominate this group with speeds of 5–9 fpm, whereas real escalators run at
90–120 fpm — these values are almost certainly missing a digit (5 → 50 fpm,
or 9 → 90 fpm). All speeds below 10 fpm are set to missing.

### Too fast

```{r}
#| label: speed-fast
elev |>
  filter(!is.na(dv_speed_fpm), dv_speed_fpm > 2000) |>
  select(dv_speed_fpm, dv_travel_distance, dv_floor_from, dv_floor_to, device_type) |>
  arrange(desc(dv_speed_fpm))
```

27 devices record speeds above 2,000 fpm. The fastest elevator installed in
NYC by 2015 (One World Trade Center) ran at approximately 2,000 fpm. Values
above 2,000 fpm are plausible for a short window of installations. The 12
records above 3,000 fpm (up to 7,000 fpm) are not: the world record in 2015
was ~4,034 fpm and was set in China, not New York. The two 6,819 fpm records
serve just 1–2 floors — a physically impossible combination of speed and
floor span. All speeds above 3,000 fpm are set to missing.

```{r}
#| label: speed-fast-zoom
elev |>
  filter(!is.na(dv_speed_fpm), dv_speed_fpm > 2000) |>
  ggplot(aes(dv_speed_fpm)) +
  geom_dotplot(binwidth = 100) +
  labs(x = "Speed (fpm)", y = NULL, title = "Speeds above 2,000 fpm")
```

There is a visible gap between the 1,400–2,000 fpm cluster (plausible
high-speed elevators) and the 3,350–7,000 fpm outliers. 3,000 fpm is a
conservative threshold that preserves the plausible high-speed devices.

## Travel distance

```{r}
#| label: dist-dist
#| fig-height: 3
elev |>
  filter(!is.na(dv_travel_distance)) |>
  ggplot(aes(dv_travel_distance)) +
  geom_histogram(binwidth = 10, boundary = 0) +
  labs(x = "Travel distance (ft)", y = "Devices")
```

```{r}
#| label: dist-small
elev |>
  filter(!is.na(dv_travel_distance), dv_travel_distance < 5) |>
  select(dv_travel_distance, dv_speed_fpm, dv_floor_from, dv_floor_to, device_type) |>
  arrange(dv_travel_distance)
```

15 records have travel distances below 5 ft. The smallest (0.33 ft, 0.43 ft,
0.62 ft) are far too short to represent any real device: even a dumbwaiter
serving just two adjacent floors would travel at least 8–10 ft. Values below
3 ft are set to missing.

## Distance vs floor span

For records with clean, purely-numeric floor pairs the implied height per
floor (travel distance / number of floors spanned) should be roughly 10–14 ft
for standard construction.

```{r}
#| label: ft-per-floor
num_floors <- elev |>
  filter(
    !is.na(dv_travel_distance),
    str_detect(coalesce(dv_floor_from, ""), "^\\d+$"),
    str_detect(coalesce(dv_floor_to, ""), "^\\d+$")
  ) |>
  mutate(
    from_n = as.integer(dv_floor_from),
    to_n   = as.integer(dv_floor_to),
    span   = to_n - from_n,
    ft_per_floor = dv_travel_distance / span
  ) |>
  filter(span > 0)

ggplot(num_floors, aes(ft_per_floor)) +
  geom_histogram(bins = 60) +
  scale_x_log10() +
  labs(x = "Implied ft per floor (log scale)", y = "Count")
```

```{r}
#| label: ft-per-floor-outliers
num_floors |>
  filter(ft_per_floor < 5 | ft_per_floor > 25) |>
  select(dv_travel_distance, dv_speed_fpm, from_n, to_n, span, ft_per_floor) |>
  arrange(ft_per_floor) |>
  head(20)
```

The outliers on the low end (< 5 ft/floor) are records where a large floor
span is paired with a small travel distance — typically a travel distance
entry of 1–4 ft for a 3–16 floor span. On the high end (> 25 ft/floor),
the pattern is a single-floor span with an implausibly large travel distance
(100–435 ft for one floor). These are travel-distance errors, not floor
errors; the floor pairs look correct. Flagging these as `dv_travel_distance`
problems:

```{r}
num_floors |>
  summarise(
    n_low  = sum(ft_per_floor < 5),
    n_high = sum(ft_per_floor > 25)
  )
```

## Travel time

For records with both a speed and a travel distance the implied travel time
in minutes is `dv_travel_distance / dv_speed_fpm`.

```{r}
#| label: travel-time-dist
#| fig-height: 3
elev |>
  filter(!is.na(dv_travel_distance), !is.na(dv_speed_fpm), dv_speed_fpm > 0) |>
  mutate(travel_min = dv_travel_distance / dv_speed_fpm) |>
  ggplot(aes(travel_min)) +
  geom_histogram(bins = 60) +
  scale_x_log10() +
  labs(x = "Implied travel time (minutes, log scale)", y = "Count")
```

The bulk of devices have travel times of a few seconds to a couple of minutes,
which is realistic. The tails hold two distinct problems.

### Implausibly short (< 3 seconds)

```{r}
#| label: travel-short
elev |>
  filter(!is.na(dv_travel_distance), !is.na(dv_speed_fpm), dv_speed_fpm > 0) |>
  mutate(travel_sec = dv_travel_distance / dv_speed_fpm * 60) |>
  filter(travel_sec < 3) |>
  select(dv_travel_distance, dv_speed_fpm, travel_sec, dv_floor_from, dv_floor_to, device_type) |>
  arrange(travel_sec)
```

Travel times below 3 seconds are physically impossible: even a brief
elevator trip requires at least 5–10 seconds to accelerate, run, and
decelerate. Every record here has either a tiny travel distance (already
flagged above, < 3 ft) or a speed above 3,000 fpm (also already flagged).
The two rules proposed above fully cover this tail.

### Implausibly long (> 10 minutes)

```{r}
#| label: travel-long
elev |>
  filter(!is.na(dv_travel_distance), !is.na(dv_speed_fpm), dv_speed_fpm > 0) |>
  mutate(travel_min = dv_travel_distance / dv_speed_fpm) |>
  filter(travel_min > 10) |>
  select(dv_travel_distance, dv_speed_fpm, travel_min, dv_floor_from, dv_floor_to, device_type) |>
  arrange(desc(travel_min))
```

The extreme cases (300–400 minutes) have speeds of 0.35 and 0.8 fpm paired
with plausible travel distances — the speed value is clearly wrong. All of
these devices have speeds already flagged as below 10 fpm. The remaining
records with travel times of 10–30 minutes are mostly escalators with speeds
of 5–8 fpm (also flagged). The > 10 min tail is fully covered by the
speed < 10 fpm rule.

## Proposed cleaning rules

The three independent rules above cover all the implausible combinations:

```{r}
#| label: rule-counts
tibble(
  rule = c(
    "dv_speed_fpm < 10 → NA",
    "dv_speed_fpm > 3000 → NA",
    "dv_travel_distance < 3 → NA"
  ),
  n_affected = c(
    sum(elev$dv_speed_fpm < 10, na.rm = TRUE),
    sum(elev$dv_speed_fpm > 3000, na.rm = TRUE),
    sum(elev$dv_travel_distance < 3, na.rm = TRUE)
  )
)
```

All three rules are implemented in `clean.R`. The travel-time tails and the
ft-per-floor outliers in the distance-vs-floor-span section are fully
explained by the speed and distance rules above; no additional floor-pair
rules are needed here beyond those already applied in `floor.qmd`.
