---
title: "Cleaning `dv_floor_from` and `dv_floor_to`"
format:
  html:
    embed-resources: true
execute:
  echo: true
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
---

```{r}
#| label: setup
#| message: false
library(dplyr)
library(stringr)
library(ggplot2)
```

The raw `DV_FLOOR_FROM` and `DV_FLOOR_TO` columns record the lowest and highest
floors served by each device. This document examines the raw values, identifies
groups and problems, and proposes the cleaning rules applied by `clean.R`.

## The raw values

```{r}
#| label: read-raw
raw <- read.csv("elevators.csv", skip = 1)
raw <- raw[!is.na(raw$BIN), ]

normalize <- function(x) {
  x <- trimws(gsub("\\s+", " ", x))
  ifelse(x == "", NA, x)
}
from_raw <- normalize(raw$DV_FLOOR_FROM)
to_raw   <- normalize(raw$DV_FLOOR_TO)

source("clean-travel-distance.R")
dist_ft <- parse_travel_distance(normalize(raw$DV_TRAVEL_DISTANCE))

tibble(
  column    = c("DV_FLOOR_FROM", "DV_FLOOR_TO"),
  n         = c(length(from_raw), length(to_raw)),
  missing   = c(sum(is.na(from_raw)), sum(is.na(to_raw))),
  pct_miss  = round(100 * missing / n),
  n_distinct = c(n_distinct(from_raw, na.rm = TRUE), n_distinct(to_raw, na.rm = TRUE))
)
```

```{r}
#| label: top-from
sort(table(from_raw), decreasing = TRUE) |> head(20)
```

```{r}
#| label: top-to
sort(table(to_raw), decreasing = TRUE) |> head(20)
```

The most common `floor_from` values are `B` (basement), `1`, `C` (cellar), and
`L` (lobby). The most common `floor_to` is `6`, reflecting that the majority
of NYC elevator devices serve low-rise residential buildings. `PH` (penthouse)
and `ST` (street) also appear.

## Classifying patterns

```{r}
#| label: classify
floor_type <- function(x) {
  case_when(
    is.na(x)                            ~ "missing",
    str_detect(x, "^\\.\\.")            ~ "dotdot prefix",
    str_detect(x, "^-?\\d+$")          ~ "integer",
    str_detect(x, "^\\d*\\.\\d+$")     ~ "decimal",
    str_detect(x, ",")                  ~ "comma-separated",
    str_detect(x, "^[A-Za-z]+$")       ~ "alpha label",
    TRUE                                ~ "other"
  )
}

all_vals <- tibble(
  value  = c(from_raw, to_raw),
  column = rep(c("floor_from", "floor_to"), each = length(from_raw))
)

all_vals |>
  mutate(type = floor_type(value)) |>
  count(type, sort = TRUE) |>
  mutate(pct = round(100 * n / sum(n), 1))
```

## Per-group findings

### Integer floors

The majority of numeric floor labels are plausible single integers:

```{r}
#| label: integer-dist
int_vals <- all_vals |>
  filter(floor_type(value) == "integer") |>
  mutate(n = as.integer(value))

summary(int_vals$n)
```

```{r}
#| label: integer-hist
ggplot(int_vals, aes(n)) +
  geom_histogram(binwidth = 5, boundary = 0) +
  labs(x = "Floor number", y = "Count")
```

The bulk of integer floors fall below 100, which is expected for NYC buildings.
Values above the floor count of NYC's tallest buildings (~104 floors as of this
data's vintage) are almost certainly data entry errors — likely elevator
specifications entered in non-floor units (e.g., a speed or capacity number
placed in the wrong field):

```{r}
#| label: integer-large
sort(unique(int_vals$n[int_vals$n > 104]))
```

```{r}
#| label: large-floor-context
tibble(
  from = from_raw,
  to   = to_raw,
  dist_ft
) |>
  filter(
    !is.na(from), !is.na(to),
    str_detect(from, "^\\d+$") | str_detect(to, "^\\d+$")
  ) |>
  mutate(
    from_n = suppressWarnings(as.integer(from)),
    to_n   = suppressWarnings(as.integer(to))
  ) |>
  filter(coalesce(from_n, 0L) > 104 | coalesce(to_n, 0L) > 104) |>
  count(from, to, dist_ft, sort = TRUE) |>
  head(20)
```

These large values appear in recurring clusters (100→300, 150→200, 100→150)
with travel distances of 10–90 ft, yielding an implied 0.2–0.5 ft per "floor"
— physically impossible. All integer floor labels above 104 are set to `NA`.

### Alpha labels

```{r}
#| label: alpha-labels
alpha_vals <- all_vals |>
  filter(floor_type(value) == "alpha label") |>
  count(value, sort = TRUE)

alpha_vals |> head(25)
```

Standard elevator designations: `B` (basement), `C` (cellar), `L` (lobby),
`G` (ground), `M` (mezzanine), `PH` (penthouse), `R` (roof), `T` (top),
`ST` (street), `SB` (sub-basement), `SC` (sub-cellar), `LL` (lower lobby).
These are all legitimate and kept as-is.

### Dotdot prefix (`..B`, `..1`, `..L`, …)

```{r}
#| label: dotdot
dotdot_vals <- all_vals |>
  filter(floor_type(value) == "dotdot prefix") |>
  mutate(stripped = str_remove(value, "^\\.+")) |>
  count(value, stripped, sort = TRUE)

dotdot_vals
```

Every dotdot value is a standard floor label with a leading `..` (occasionally
`.`). Comparing the stripped label against the non-dotdot counterparts:

```{r}
#| label: dotdot-compare
# Compare the top-5 dotdot from values to their non-dotdot equivalents
tibble(from = from_raw, to = to_raw) |>
  filter(from %in% c("..B", "..1", "..L", "..C")) |>
  count(from, to, sort = TRUE) |>
  head(20)
```

```{r}
tibble(from = from_raw, to = to_raw) |>
  filter(from %in% c("B", "1", "L", "C")) |>
  count(from, to, sort = TRUE) |>
  head(20)
```

The `..B → 6` pattern (102 cases) mirrors exactly the `B → 6` pattern (6,775
cases). The leading dots carry no information and are stripped: `..B` → `B`,
`..1` → `1`, etc.

### Decimal floors

```{r}
#| label: decimal
decimal_vals <- all_vals |>
  filter(floor_type(value) == "decimal") |>
  count(value, sort = TRUE)

decimal_vals
```

Values like `0.1`, `0.2`, `0.6` are not recognizable floor designations in
any standard NYC building convention, and almost all records containing them
have no travel distance:

```{r}
#| label: decimal-context
tibble(from = from_raw, to = to_raw, dist_ft) |>
  filter(
    str_detect(coalesce(from, ""), "^\\d*\\.\\d+$") |
    str_detect(coalesce(to, ""),   "^\\d*\\.\\d+$")
  ) |>
  summarise(
    n = n(),
    n_with_distance = sum(!is.na(dist_ft))
  )
```

These are most likely data entry errors (perhaps fractional numbers intended
to represent something else). All decimal floor values are set to `NA`.

### Comma-separated floors

```{r}
#| label: comma
all_vals |>
  filter(floor_type(value) == "comma-separated") |>
  count(value, sort = TRUE) |>
  head(20)
```

Values like `B,1` or `C,1` appear in `floor_from` and describe elevators with
two entrance points at the bottom (e.g., an entrance at both the cellar and
the ground floor). These multi-level lobby arrangements are common in NYC
buildings. The values are kept as-is; they are not parseable to a single
integer but are valid labels.

## Cross-column consistency

### Inverted floors

For purely numeric floor pairs, `floor_from` should be ≤ `floor_to`. Checking:

```{r}
#| label: inverted
tibble(from = from_raw, to = to_raw) |>
  filter(
    str_detect(coalesce(from, ""), "^\\d+$"),
    str_detect(coalesce(to, ""), "^\\d+$")
  ) |>
  mutate(from_n = as.integer(from), to_n = as.integer(to)) |>
  summarise(
    n_inverted = sum(from_n > to_n),
    n_equal    = sum(from_n == to_n),
    n_normal   = sum(from_n < to_n)
  )
```

141 records have `floor_from > floor_to` numerically. These are most likely
data entry transpositions rather than deliberate encoding (there is no
convention where a higher numeric `floor_from` is meaningful):

```{r}
#| label: inverted-examples
tibble(from = from_raw, to = to_raw, dist_ft) |>
  filter(
    str_detect(coalesce(from, ""), "^\\d+$"),
    str_detect(coalesce(to, ""), "^\\d+$"),
    as.integer(from) > as.integer(to)
  ) |>
  arrange(desc(as.integer(from) - as.integer(to))) |>
  head(10)
```

The largest inversions (e.g., 11→1, 6→3) look like transpositions. These are
corrected by swapping `floor_from` and `floor_to`.

### Travel distance consistency

For records with numeric floor pairs and a recorded travel distance, the
implied feet per floor should be roughly 10–14 (standard floor-to-ceiling
height):

```{r}
#| label: ft-per-floor
tibble(from = from_raw, to = to_raw, dist_ft) |>
  filter(
    str_detect(coalesce(from, ""), "^\\d+$"),
    str_detect(coalesce(to, ""), "^\\d+$"),
    !is.na(dist_ft)
  ) |>
  mutate(
    from_n = as.integer(from),
    to_n   = as.integer(to),
    span   = abs(to_n - from_n),
    ft_per_floor = dist_ft / span
  ) |>
  filter(span > 0, !is.na(ft_per_floor)) |>
  ggplot(aes(ft_per_floor)) +
  geom_histogram(bins = 60) +
  scale_x_log10() +
  labs(x = "Implied ft per floor (log scale)", y = "Count")
```

The cluster around 10 ft/floor is reassuring. The left tail (< 2 ft/floor)
and right tail (> 100 ft/floor) contain the anomalies already identified:
large integer floor numbers and single-floor spans with implausibly large
travel distances.

## Summary of cleaning rules

| Problem | Rule |
|---|---|
| `..B`, `..1`, etc. | Strip leading dots |
| Decimal floors (`0.1`, `0.6`, …) | Set to `NA` |
| Integer floor > 104 | Set to `NA` |
| `floor_from` > `floor_to` (both integers) | Swap the two values |
