---
title: "Cleaning `dv_travel_distance`"
format:
  html:
    embed-resources: true
execute:
  echo: true
knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
---

```{r}
#| label: setup
#| message: false
library(dplyr)
library(stringr)
library(ggplot2)
```

The raw `DV_TRAVEL_DISTANCE` column records the vertical travel distance of
each device as free text, with no consistent units or format. This document
looks at the raw values, divides them into groups, works out what each group
means, and then derives the parsing rules implemented by
`parse_travel_distance()` in `clean-travel-distance.R`, which `clean.R` uses
to convert the column to numeric feet.

## The raw values

We read the raw strings directly from the CSV, applying the same whitespace
normalization as `clean.R`:

```{r}
#| label: read-raw
raw <- read.csv("elevators.csv", skip = 1)
raw <- raw[!is.na(raw$BIN), ]
distance <- trimws(gsub("\\s+", " ", raw$DV_TRAVEL_DISTANCE))
distance[distance == ""] <- NA

tibble(
  n = length(distance),
  missing = sum(is.na(distance)),
  pct_missing = round(100 * mean(is.na(distance))),
  unique_values = n_distinct(distance, na.rm = TRUE)
)
```

About 62% of devices have no recorded travel distance, and the ~29k recorded
values use ~4,800 distinct strings. The most common values hint at the mess:

```{r}
#| label: top-values
sort(table(distance), decreasing = TRUE) |> head(20)
```

`60`, `60'`, and `60 FT` all clearly mean the same thing.

## Dividing into groups

Classifying every value by its lexical pattern:

```{r}
#| label: groups
pattern_group <- function(x) {
  case_when(
    is.na(x) ~ "missing",
    str_detect(x, "^\\d{4}-\\d{2}-\\d{2}$") ~ "date",
    str_detect(x, "^\\d+$") ~ "bare number",
    str_detect(x, "^\\d+\\s*'$") ~ "feet with '",
    str_detect(x, "^\\d+\\s*'\\s*\\d+\\s*\"?$") ~ "feet ' inches",
    str_detect(x, "^\\d+\\s*-\\s*\\d+\\s*\"?$") ~ "feet-inches dash",
    str_detect(x, regex("FT|FEET", ignore_case = TRUE)) ~ "explicit FT",
    str_detect(x, "^\\d+\\s*\"$") ~ "trailing \" only",
    str_detect(x, "^\\d*\\.\\d+$") ~ "decimal",
    TRUE ~ "other"
  )
}
grp <- pattern_group(distance)

tibble(value = distance, group = grp) |>
  filter(!is.na(value)) |>
  summarise(
    n = n(),
    distinct = n_distinct(value),
    examples = paste(head(unique(value), 5), collapse = ", "),
    .by = group
  ) |>
  arrange(desc(n))
```

## Per-group findings

### Bare numbers (~9.4k values)

The bulk of the column. The distribution (median 80, 99th percentile ~635)
matches travel measured in **feet**: typical NYC elevators travel tens to a
few hundred feet. A handful of values are physically impossible:

```{r}
#| label: bare-numbers
bare <- as.numeric(distance[grp == "bare number"])
summary(bare)
sort(unique(bare[bare > 1300]))
```

The tallest elevator travel in NYC at the time of this data (late 2015) is
roughly 1,250–1,300 ft (One World Trade Center), so values above 1,300 —
and a single `0` — are set to missing. Everything else is accepted as feet.

### Feet with a quote mark, or an explicit unit (~9.1k values)

`60'`, `24FT`, `99 FT`, `6.5FT` are unambiguous feet. A few oddballs like
`35FT"` and `39"FT` carry stray quote marks but the explicit `FT` settles
the unit. All are accepted as feet.

### Feet and inches (~8.8k values)

The second-largest family writes feet and inches with a variety of
separators:

* apostrophe: `16'4"`, `103'1`, `14'0'`
* dash: `73-10`, `60'-4`, `9-11'`
* comma: `17,8`, `20,10`
* space: `307 3`, `18 6"`
* mistyped symbols: `33"6"`, `48"11`, `45"3"` (a `"` where the `'` belongs)

The giveaway that the second number is inches: across all two-number values
it is almost always 0–11.

```{r}
#| label: inches-distribution
two_num <- str_match(distance, "^(\\d+)\\s*['\",-]\\s*-?(\\d{1,2})[\"'.:]*$")
table(as.numeric(two_num[, 3]))
```

Values where the second number is 12 or more (`13'13`, `19'91`, `90/77`,
about 50 in total) can't be feet-and-inches, and we have no way to know what
was meant, so they are set to missing. The rest are converted as
`feet + inches / 12`.

We do *not* treat `/` as a feet-inches separator: values like `331/2`,
`537/8`, and `141/2` read more naturally as mixed fractions (33½, 53⅞,
14½), so slash-separated values are ambiguous and set to missing (~30
values).

### Decimals (~880 values)

Values like `116.6`, `51.5`, `19.79`. Two readings are possible — decimal
feet, or feet.inches — but the two-digit fractional parts (`.79`, `.92`)
rule out inches for many, and for single-digit fractions the two readings
differ by at most half a foot. Accepted as decimal feet.

### Dates: Excel mangling (33 values)

A surprise group: values like `2015-11-04`. These are feet-and-inches
values (`11-4`) that Excel helpfully converted to dates in the
then-current year when the FOIL extract was prepared (the data was received
in November 2015). The evidence is strong: 31 of 33 have year 2015, and the
"day" part is never above 11 — exactly what you'd expect if it were inches,
and very unlikely for real dates.

```{r}
#| label: dates
table(distance[grp == "date"])
```

The 2015 dates are recovered as `month + day / 12` feet. The two stray
years (`2000-02-01`, `2004-04-03`) don't fit the pattern and are set to
missing.

### Trailing `"` only: genuinely ambiguous (~130 values)

Values like `188"` or `1858"` look like inches, but cross-checking against
the floors served shows the group is a mix of mistyped feet and true
inches:

```{r}
#| label: ambiguous-inches
tibble(
  value = distance,
  from = raw$DV_FLOOR_FROM,
  to = raw$DV_FLOOR_TO
) |>
  filter(value %in% c("188\"", "1858\"", "552\"", "21\"", "102\"")) |>
  distinct()
```

* `188"` serving L,1→21 must be 188 **feet** (188 in ≈ 16 ft can't span 21
  floors).
* `1858"` serving C→13 must be 1,858 **inches** (≈ 155 ft; 1,858 ft is
  impossible).
* `552"` serving 1→5 must be inches (≈ 46 ft); `21"` on a dumbwaiter
  serving 1→3 must be feet.

Since the unit flips value-by-value, no rule is reliable. All
single-number values whose only annotation is a trailing `"` are set to
missing.

### Junk (~100 values)

Letter codes and symbols with no recoverable number: `E`, `B`, `EXIST`,
`ABOVE`, `BASE`, `-----`, `(E)`, `{`, `` ` ``. All set to missing.

## The cleaning function

The rules above are implemented in `parse_travel_distance()`, defined in
`clean-travel-distance.R` and sourced by `clean.R`, which uses it to convert
`dv_travel_distance` to numeric feet in `elevators.parquet`:

```{r}
#| label: source-clean
source("clean-travel-distance.R")
parse_travel_distance
```

## Results

```{r}
#| label: apply
parsed <- parse_travel_distance(distance)

tibble(value = distance, group = grp, parsed = parsed) |>
  filter(!is.na(value)) |>
  summarise(n = n(), parsed = sum(!is.na(parsed)), .by = group) |>
  mutate(pct_parsed = round(100 * parsed / n)) |>
  arrange(desc(n))
```

Overall `r round(100 * sum(!is.na(parsed)) / sum(!is.na(distance)), 1)`% of
non-missing raw values were converted; the remaining
`r sum(!is.na(distance) & is.na(parsed))` were rejected as ambiguous,
impossible, or junk. The resulting distribution is plausible: most devices
travel 10–200 ft, with a long tail of skyscraper elevators.

```{r}
#| label: histogram
#| fig-height: 3
ggplot(tibble(ft = parsed), aes(ft)) +
  geom_histogram(binwidth = 10, na.rm = TRUE) +
  scale_x_continuous(limits = c(0, 800)) +
  labs(x = "Travel distance (ft)", y = "Devices")
```

```{r}
#| label: summary
summary(parsed)
```
